---
title: "DSC595_001_A2_your_ncsu_email_prefix(mine is jshouse)"
author: "Your First and Last Name"
date: "Due 5pm on Friday, January 26th"
output:
  html_document:
    code_folding: show
    toc: yes
    toc_depth: 2
    toc_float: yes
---

# Assignment 2

#### For all assignments, unless otherwise stated, your work will always consist of two files:
- The .RMD file with the code (this file) 
- The .HTML file resulting from you `knitting` this file 
<br><br><br>
 
### For this Assignment, you will be working on some content that we have not gone over in class and will go over the next time we meet 

### Review of Syntax 
<br> 

#### When I use syntax that looks like this `john()` it means that `john` is a function and the `john()` is more explicitly drawing your attention to the fact because of the `()`.  

- Remember in `R`, all function calls use the `()` after the function
- Inside the `()` is where you pass parameter values to the function 
  - Thus last week, `rnorm()` was a function call and we passed the value of `100` for the parameter `n` when we called the function
  - `rnorm(n=100)` then called the function with the passed value of the parameter `n` as 100.
  - if we type ?rnorm at the console prompt, we can see that this function actually has 3 parameters `rnorm(n, mean = 0, sd = 1)`
    - `n` : This parameter does **NOT** have a default value, so it **MUST** be supplied when the function is called
    - `mean = 0` : This parameter does have a default value, and will use the value of `0` unless you change it 
    - `sd = 1` : This parameter does have a default value, and will use the value of `1` unless you change it 
  - Thus when we type `rnorm(100)`, R assumes we want `rnorm(n = 100, mean = 0, sd = 1)`


## Exercise 1  

- Part 1: Use the `rnorm()` function to create 3 objects in memory named as you see fit
  - each one should have n = 100
  - each one should have a different values passed to the `sd` and `mean` parameters 
  <br>
- Part 2: Plot a histogram of each object (3 graphs) and title each one with information regarding your choice of mean and sd 
  - use the `hist()` function to make the plots  
<br>
- Part 3: Explain why every time you submit the same `rnorm()` function call, the values are different. Tell me in words how you would make that call **reproducible** so that someone else would get the same values for the same function call.  

Part 3 Answere: type here 

```{r}
# Type code here to answer Parts 1 & 2

```

# R Data Types 

#### R has many different types of objects. You can interrogate and get information about an R object using the `str()` function in R. `str` stands for structure.  The syntax is `str(object)` 

## Exercise 2

- Part 1: You created 3 objects in Exercise 1 using the `rnorm()` function. Determine information about that object and print to screen 
- Part 2: Do the same with the iris dataset used last week 
- Part 3: Which variable in the iris dataset is of the type `factor`?

Part 3 Answer:  

```{r}
# Type code here to answer parts 1 & 2

```


## Exercise 3 

I've created an object named `myvec` 

- Part 1: Have R tell you how many elements are in that vector
- Part 2: Create a new object named `myvec_subset` where you subset `myvec` to the 1st, 10th, and 13th elements and print that new object 
- Part 3: Create a new vector that appends a 27th element to `myvec` with the value of "Capital Letters!"  and print the new object 

```{r}
myvec <- LETTERS
# continue with code in this chunk to answer parts 1,2, and 3

```

## Exercise 4 

#### Use the built-in `iris` dataset 

Part 1: Use `plot()` to plot a scatterplot where 
 - X is the Sepal.Length 
 - Y is the Sepal.Width
 - the points are solid again
 - the points are colored by species
 - title the plot to describe it
 
Part 2: Subset the iris dataset and make the same scatterplot but only for the Species "setosa"
- title appropriately
 
Part 3: instead of specifying columns for a single plot
  - use `plot(iris)` 
  - color by species and make solid again 

```{r message=F}
## look at first 6 records of iris dataset 
head(iris)
# Answer parts 1,2, and 3 by typing code here that will display results when you knit an html

```


  